Diversification of Australian blind snakes

### 2022-11-16 manuscript.Rmd
rewrote some parts of intro, included more info on Biogeobears and GeoHiSSE methods.

Present-day distrubtion of extant species offers clues about their biogeographic history and how the lineages may have diversified.

The aridification on continental Australia over the past 20 million years resulted in a diversity of habitats and spurred multiple radiations of flora and fauna across the continent, many of which are endemic [@byrneBirth2008;@melvilleEvolutionary2011]. The arid zone makes up the majority of the Australian landscape and provides habitat to ~43% of all Australia’s terrestrial squamate reptiles (lizards and snakes). Reptiles represent the largest vertebrate diversity in Australia with over 1000 species (~800 lizards and ~200 snakes), many of which diversified throughout the Miocene (starting ~25 MYA) to the present [@oliverPhylogenetic2017]. For example, agamid dragons [@chenSexual2012], monitor lizards [@brennanPhylogenomics2021], _Sphenomorphus_ and _Eugongylus_ skinks [@skinnerLygosomine2011], _Gehyra_ geckos [@ashmanDiversification2018], elapid snakes [@leeDiversification2016], pythons [@esquerrePhylogenomics2020], and blind snakes [@marinTracing2013] are reptiles that colonised Australia from elsewhere and have diversified remarkably from the beginning of the Miocene.

Roughly a quarter of the snake species diversity in Australia is represented by blindsnakes of the genus _Anilios_, which comprises **47** described species [@uetzQuarter2021]. All _Anilios_ species are found in Australia except for _A. eryncinus_, which is only found in Papua New Guinea. While morphologically conserved and difficult to distinguished, extensive phylogenetic studies and several efforts to revisit specimens revealed the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@ellisTyphlopid2019]. The diversification is reported to have originated from a single colonisation event by a non-arid adapted ancestor from Asia between 19-39 mya [@vidalBlindsnake2010;@marinTracing2013], before the onset of aridification over the past 20 million years [@byrneBirth2008].

However, the study did not account for alternative biogeographic areas (e.g. WWF Ecoregions) and models, or uncertainty in phylogenetic relationships and divergence times. Despite representing a substantial portion of snake biodiversity, there has been few investigations into the macroevolutionary dynamics of blind snakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic ecological changes because they are geographically restricted to the continent and can be found in almost every biome.

@@

## Biogeography

To estimate the ancestral biome of blind snakes in Australia, we used the R package `BioGeoBEARS` 1.1.1 [@matzkeBioGeoBEARS2013]. BioGeoBEARS provides a flexible likelihood-based framework that models how the branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events including processes like dispersal, extinction, or switching to adjacent biomes) or at the time of branching (cladogenetic events involving sympatric, vicariant and founder-event speciation). The fundamental inputs for the software include a time-calibrated phylogeny, information about the biome/range each species are found, and a historical biogeographic model. We divided Australo-Papua into nine discrete biomes and assigned each Australian species to specific biome(s) based on their occurrence records as described above. Biome classification were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World dataset [@olsonTerrestrial2001] and followed the nomenclature of published Austrlia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021]. 
The phylogeny and distribution information were then applied to widely used historical biogeographical methods including likelihood implementations of Dispersal-Vicariance Analysis [@ronquistDispersalVicariance1997] (DIVALIKE), Dispersal-Extinction-Cladogenesis [@reeMaximum2008] (DECLIKE), and BayArea [@landisBayesian2013] (BAYAREALIKE) in `BioGeoBEARS` [@matzkeBioGeoBEARS2013]. We included two additional free parameters that we thought is relevant. The first parameter *j* assigns per-event weight to founder-events [@matzkeModel2014], and the second parameter _x_ modifies the probability of dispersal based on distance between biomes [@vandamEvaluating2016] (see Appendix-method-biogeobears). We used likelihood ratio tests to determine if the additional free parameter is a justified improvement over the simpler base models, and selected the model that fitted the data best based on corrected Akaike Information Criterion (AICc) scores.

@@

## Biogeographic-dependent diversification analyses

We further investigated whether broad patterns of diversification are associated with geographic range. Using the R package `hisse` v. 2.1.9 [@beaulieuDetecting2016], we fitted five biogeographic-dependent speciation-extinction (SSE) models under the GeoSSE [@goldbergPhylogenetic2011] and GeoHiSSE [@caetanoHidden2018] framework to the phylogeny and geographic range data. 
Since SSE models only allow three states, we restricted our geographic states to species that have distributions in arid areas, non-arid areas, and those that are widespread across both arid and non-arid areas. 
Consistent with our BioGeoBEARS anaylsis, we classified species that primarily occur in the arid zone as arid (n = 8), species that occur primarily outside of the arid zone as non-arid (n = 24), and those that occupy both the arid and non-arid zone as widespread (n = 5). We assumed uniform taxon sampling for 89% in arid areas, 71% in non-arid areas, and 83% in widespread areas based on number of species in the phylogeny and all currently described species of _Anilios_.
The five SSE models vary by whether they allow diversification rates to be associated with geographic range and/or with unobserved 'hidden states'. Model 1 assumes equal diversification rates regardless of geographic range, model 2 allows geographic range to affect diversification rates (canonical GeoSSE model). Model 3 and 4 are derivation of models 1 and 2 but with unobserved hidden states (see Table GeoHiSSE table). The last model is a complementary model that removes cladogenetic effect entirely. We compared the AIC scores to select the best models. Following the approach in @caetanoHidden2018, we averaged the estimates of diversification parameters and ancestral geographic range estimate across all models according to their AIC weights.

@@

## Statistical analyses
Statistical analyses were run in R version 4.2 (CITE R) and Julia (CITE). We visualised dynamics using baseR and `ggplot2` in R. The Australian bioregion map visualised in `QGIS` v (CITE Australian Map).

@@

>>>>

Present-day distrubtion of extant species offers clues about their biogeographic history and insight into the tempo and mode with which different groups diversified.

The aridification on continental Australia began around 20 million years ago with pockets of arid zone developing and subsequent extreme warming and cooling periods in last ~5 million years creating what is now known as the 'outback' [@byrneBirth2008]. From there arid-adapted vegetation expanded and the mesic biomes became more fragmented [@crispHow2013;@martinCenozoic2006]. This period of aridification is believed to involve rapid speciation of many arid-adapted vertebrates [@raboskyExceptional2007;@pepperPalaeoclimate2011;@brennanMass2017] and the extinction of some mesic-adapted groups [@bryantLines2016;@cardilloPhylogeny2017;@owenHow2017].

Today, the arid zone covers the majority of area in the Australian landscape (70%) and provides habitat to 43% of all of Australia's terrestrial reptiles, termites, and ants [@byrneBirth2008]. Squamate reptiles (lizards and snakes) represent the largest vertebrate diversity in Australia with over 1000 species (~800 lizards and ~200 snakes), many of which diversified in the last 20 - 6 myr [@brennanMass2017;@oliverPhylogenetic2017].

Some clades, like Elapid snakes [@leeDiversification2016;@sandersMolecular2008] and scincid lizards [[@raboskyExceptional2007] showed dramatic ecological diversity taking up different niches .

Other clades like geckos and blind snakes remain morphologically conserved and were thought to be species poor until recent molecular data suggested otherwise [@marinHidden2013;@moritzCryptic2018;@zozayaAre2019].

The species richness and distribution of squamate reptiles across different biomes makes them a good group to test hypotheses about the tempo and mode of diversification in Australia.

For example, agamid dragons [@chenSexual2012], monitor lizards [@brennanPhylogenomics2021], _Sphenomorphus_ and _Eugongylus_ skinks [@skinnerLygosomine2011], _Gehyra_ geckos [@ashmanDiversification2018], elapid snakes [@leeDiversification2016], pythons [@esquerrePhylogenomics2020], and blind snakes [@marinTracing2013] are reptiles that colonised Australia from elsewhere and have diversified remarkably from the beginning of the Miocene.

Roughly a quarter of the snake species diversity in Australia is represented by blindsnakes of the genus _Anilios_, which comprises **47** described species [@uetzQuarter2021]. All currently described species of _Anilios_ are found in Australia except for _A. eryncinus_, which is endemic to New Guinea. Blind snakes are morphologically conserved and often difficult to distinguis. Extensive phylogenetic studies and several efforts to revisit specimens revealed the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@hedgesTaxonomic2014;@ellisTyphlopid2019].

The diversification of _Anilios_ is reported to have originated from a single colonisation event by a mesic adapted ancestor from Asia between 19-39 mya [@vidalBlindsnake2010;@marinTracing2013], before the onset of aridification over the past 20 million years [@byrneBirth2008]. However, how and

However, the study did not account for alternative biogeographic areas (e.g. WWF Ecoregions) and models, or uncertainty in phylogenetic relationships and divergence times.

Despite representing a substantial portion of snake biodiversity, there has been few investigations into the macroevolutionary dynamics of blind snakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic ecological changes because they are geographically restricted to the continent and can be found in almost every biome.

@@

## Biogeography and biome transitions

To investigate how the distribution of biome evolved in blind snakes we used the R package `BioGeoBEARS` 1.1.1 [@matzkeBioGeoBEARS2013]. BioGeoBEARS provides a flexible likelihood-based framework that models how the branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events including processes like dispersal, extinction, or switching to adjacent biomes) or at the time of branching (cladogenetic events involving sympatric, vicariant and founder-event speciation). The fundamental inputs for the software include a time-calibrated phylogeny, information about the biome/range each species are found, and a historical biogeographic model. We divided Australo-Papua into nine discrete biomes and assigned each Australian species to one or more biomes based on their occurrence records. Biome classification were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World dataset [@olsonTerrestrial2001] and followed the nomenclature of published Austrlia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021]. 
The phylogeny and distribution information were then applied to widely used historical biogeographical methods including likelihood implementations of Dispersal-Vicariance Analysis [@ronquistDispersalVicariance1997] (DIVALIKE), Dispersal-Extinction-Cladogenesis [@reeMaximum2008] (DECLIKE), and BayArea [@landisBayesian2013] (BAYAREALIKE) in `BioGeoBEARS` [@matzkeBioGeoBEARS2013]. We included two additional free parameters that we thought is relevant. The first parameter *j* assigns per-event weight to founder-events [@matzkeModel2014], and the second parameter _x_ modifies the probability of dispersal based on distance between biomes [@vandamEvaluating2016] (see Appendix-method-biogeobears). 
We used likelihood ratio tests to determine if the additional free parameter is a justified improvement over the simpler base models, and selected the model that fitted the data best based on corrected Akaike Information Criterion (AICc) scores. Based on the best model we ran an additional 50 biogeographic stochastic maps to account for uncertainty in ancestral biome estimation and trasition between biomes, and to estimate the frequency and directionality of dispersal events across biomes [@mat.

@@

## Biogeographic-dependent speciation

We further investigated whether diversification rates and biome transitions differ between geographic areas. Using the R package `hisse` v. 2.1.9 [@beaulieuDetecting2016], we fitted five biogeographic-dependent speciation-extinction (SSE) models under the GeoSSE [@goldbergPhylogenetic2011] and GeoHiSSE [@caetanoHidden2018] framework to the phylogeny and geographic range data. 
Since SSE models only allow two states, we restricted our geographic states to species that have distributions in arid areas, non-arid areas, and those that are widespread across both arid and non-arid areas. 
Consistent with our BioGeoBEARS anaylsis, we classified species that primarily occur in the arid zone as arid (n = 8), species that occur primarily outside of the arid zone as non-arid (n = 24), and those that occupy both the arid and non-arid zone as widespread (n = 5). We assumed uniform taxon sampling for 89% in arid areas, 71% in non-arid areas, and 83% in widespread areas based on number of species in the phylogeny and all currently described species of _Anilios_.
The five SSE models vary by whether they allow diversification rates to be associated with geographic range and/or with unobserved 'hidden states'. Model 1 assumes equal diversification rates regardless of geographic range, model 2 allows geographic range to affect diversification rates (canonical GeoSSE model). Model 3 and 4 are derivation of models 1 and 2 but with unobserved hidden states (see Table GeoHiSSE table). The last model is a complementary model that removes cladogenetic effect entirely. We compared the AIC scores to select the best models. Following the approach in @caetanoHidden2018, we averaged the estimates of diversification parameters and ancestral geographic range estimate across all models according to their AIC weights.

@@

## Data visualisation
Base data visualisations were performed in run in R version 4.2 (CITE R) using baseR and `ggplot2` and post-processed in Adobe Illustrator. The Australian bioregion map visualised in `QGIS` v (CITE Australian Map).

@@

Microhabitat better explains diversification than climate, but typhlopidae has the second highest speciation rates among fossorial snakes [@bars-closelDiversification2017].

Future phylogeographic study should consider densely sampling species to see whether there are more expansion into or out of arid zone for widespread taxa. Two pulses of speciation early Miocene and then another early Pleistocene. Geographic barrier for mesic adapted lineages in the temperate and tropical forests may have a hard time crossing the lines [@bryantLines2016]
